siamcat_MND_MRD

### 2020-04-03 annals_manuscript.Rmd
updated

#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object. curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
data(atlas1006)
atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd))

@@

  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = wes_palette(glob_theme))

@@

@@

plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = " ", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()

@@

@@

abundance.bar(atlas.pseq, "atlas_bar")
abundance.bar(pseq.subset, "pseq_bar")

@@

NOTHING BELOW THIS WORKS!!!!!!!!

## MICROBIOME COMPOSITION

```{r Phylum #}

# Show available ranks in the dataset
rank_names(pseq)

# Create table, number of features for each phyla
table(tax_table(pseq)[, "Phylum"], exclude = NULL)

This shows a few phyla for which only one feature was observed. Those may be worth filtering.

guide_italics
ggtitle("Relative abundance") +  
```{r}

clos.ph <- aggregate_taxa(clostridium, "Phylum")

clos.ph.rel <- microbiome::transform(clos.ph, "compositional")

plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
  theme(legend.position = "bottom", ) + theme_bw() + 
  theme(axis.text.x=element_blank()) + 
  theme(legend.title = element_text(size=18), legend.text = element_text( face="italic")) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend(title="Phylum"))

print(plot.comp.rel)

ggsave("phylum_relative_abundance_clos.png")

plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none") + guide_italics
ggsave("phylum_relative_abundance_clos2.png")

```{r CLostridium Bar Plot}

clos <- aggregate_taxa(clostridium, "Phylum")

pseq.cl.rel <- microbiome::transform(clos, "compositional")

plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic"))

plot.cl.rel + scale_fill_brewer(palette = "Paired")
ggsave("cl_abundance_bar.png")

```{r Clostridium genus bar plots}

pseq.cl.rel <- microbiome::transform(clostridium, "compositional")

plot.cl.rel <- plot_composition(pseq.cl.rel, average_by = "mnd", x.label = "sample", plot.type = "barplot")

plot.cl.rel + scale_fill_viridis(discrete = TRUE, option = "A")

```{r clostridium heatmap}

pseq.c.ph <- aggregate_taxa(clostridium, "Phylum")

pseq.c.ph.rel <- microbiome::transform(pseq.c.ph, "compositional")

plot.c.comp.rel <- plot_composition(pseq.c.ph.rel, x.label = "sample", plot.type = "heatmap")

p <- plot.c.comp.rel + coord_flip()
p + theme(axis.text.x=element_blank()) + labs(y = "Sample")
ggsave("clos_histogram.png")

## Community composition plotting
Classic bar plots of bacterial phyla present per sample can be useful for communicating "high level" results. These are relatively easy to interpret when major shifts in microbial communities are present, such as in this study where antibiotics are used However, they are not effective at detecting subtle shifts in communities or taxa and do not convey any statistical significance and can be subjectively interpreted. Interpretation of these plots should always be subject to subsequent statistical analysis.

```{r}

clos <- aggregate_taxa(clostridium, "Phylum")

# agglomerate at phylum level
clos_phy <- tax_glom(clos, taxrank = "Phylum")

# Transform to rel. abundance (or use ps0.ra)
clos_phyl <- transform_sample_counts(clos_phy, function(x) {x/sum(x)})

# Melt to long format for easy ggploting
clos_phylum <- psmelt(clos_phyl)

# Plot - Phylum
p.ra.phylum <- ggplot(clos_phylum, aes(x = sample, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~mnd, scales = "free_x", nrow = 4, ncol = 7) +
  theme(axis.text.x = element_blank()) +
  theme(axis.title.x = element_blank()) +
  labs(title = "Abundant Phylum of Clostridium(> 1%)")

p <- p.ra.phylum + scale_fill_viridis(discrete = TRUE, option = "A")
plot(p)
ggsave("phylum_abun_mnd_clos.png")

```{r}

get_taxa_unique(pseq, "Phylum")

```{r}

get_taxa_unique(pseq, "Genus")

```{r}

get_taxa_unique(clostridium, "Genus")

```{r}

df <- psmelt(pseq)
kable(head(df))

```{r}

head(abundances(pseq)["Faecalibacterium prausnitzii et rel.",])

>>>>

#For the MetaPhlAn2 bugs datasets (but not other data types), you gain a lot of taxonomy-aware, ecological analysis and plotting by conversion to a phyloseq class object. curatedMetagenomicData provides the ExpressionSet2phyloseq() function to make this easy:
data(atlas1006)
atlas.pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006
atlas.pseq <- subset_samples(atlas.pseq, !is.na(mnd), !is.na(nationality) )

@@

  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))

@@

#Abundance plots

@@

plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()

@@

# atlas barplot
p.new <- aggregate_taxa(atlas.pseq, "Phylum")

plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()

p <- plot.p.new.rel + scale_fill_brewer(palette = "Paired")
ggsave("atlas_bar.png")

# pseq barplot
p.new <- aggregate_taxa(pseq.subset, "Phylum")

plot.p.new.rel <- plot_composition(p.new.rel, average_by = "mnd", x.label = "MND DALYs", plot.type = "barplot") + guides(fill=guide_legend(title="Phylum")) + theme(legend.text = element_text( face="italic")) + theme_classic()

p1 <- plot.p.new.rel + scale_fill_manual(values=c( "#A5CEE3", "#1F78B4", "#34A02C", "#E3211D", "#FF7E00", "blue", "yellow", "red"))
ggsave("pseq_bar.png")

p3 <<-plot_grid(title,p,p1, align = "v" , ncol = 1, labels = c('','A', 'B'),rel_heights = c(0.1, 1,1.2))
plot(p3)

@@

abundance.bar(atlas.pseq, "atlas_bar")
#abundance.bar.pseq(pseq.subset, "pseq_bar")

# Alpha and beta diversity figure
```{r}

# gets the meta data 
  pseq.meta <- meta(atlas.pseq)
  # generates the diversites
  tab <- microbiome::alpha(atlas.pseq, index = "shannon")

  # Compare differences in Shannon index between mnd group of the study subjects
  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
  # on to the graphs!!!

  # This country versus nationality
  p1<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
                  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")

  # the number in the palettes must equal the number of groups in bmi.pairs

  p2 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))

# gets the meta data 
  pseq.meta <- meta(pseq.subset)
  # generates the diversites
  tab <- microbiome::alpha(pseq.subset, index = "shannon")

  # Compare differences in Shannon index between mnd group of the study subjects
  pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")
  # on to the graphs!!!

  # This country versus nationality
  p3<-   ggviolin(pseq.meta, x = "nationality", y = "Shannon",
                  add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")

  # the number in the palettes must equal the number of groups in bmi.pairs

  p4 <<- ggviolin(data=subset(pseq.meta, !is.na(mnd)), x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", na.rm = TRUE, palette = c("yellow", "orange", "red"))

  p4 <<- p4 + stat_compare_means(comparisons = mnd.pairs) + xlab("MND DALYs") + theme(legend.position="none")

p.1 <<-plot_grid(title, p1, p3, p2, p4, align = "v" , ncol = 2, labels = c('','A', 'B', 'C', 'D'), rel_widths = 1, rel_heights = 2)
print(p.1)

@@

### 2020-04-03 siamcat_MND_MRD.Rmd
updated

```{r}
pseq.c <- microbiome::transform(pseq.1, "compositional")
```

@@

label <- create.label(meta=sample_data(pseq.c),
    label = "mnd",
    case = "HIGH", control = "LOW")

#FIRST WAY:
# siamcat <- siamcat(feat=feat, label=label, meta=meta)
# siamcat <- siamcat(feat=otu_table, label=label, meta=sample_data(pseq))
# Error in validObject(.Object) : invalid class “phyloseq” object: 
 # An otu_table is required for most analysis / graphics in the phyloseq-package

#SECOND WAY
# siamcat <- siamcat(phyloseq=phyloseq, label=label)
siamcat <- siamcat(phyloseq=pseq.c, label=label)

@@

## Feature Filtering
Now, we can filter feature with low overall abundance and prevalence.
```{r}

@@

Since we have quite a lot of microbial species in the dataset at the moment, we can perform unsupervised feature selection using the function filter.features.
```{r}

siamcat <- filter.features(siamcat,
    filter.method = 'abundance',
    cutoff = 0.001)

@@

check.confounders(siamcat,
    fn.plot = 'conf_checkMND.pdf', meta.in = c("mnd", "age", "gender", "BMI", "age_category", "born_method", "feeding_practice"))

#### Association Testing ####

siamcat <- check.associations(
    siamcat,
    sort.by = 'fc',
    fn.plot = 'assocMND.pdf',
    alpha = 0.05,
    mult.corr = "fdr",
    detect.lim = 10 ^-6,
    plot.type = "quantile.box",
    panels = c("fc", "prevalence", "auroc"))

@@

@@

library(randomForest)

@@

#### Interpretation Plot

@@

model.interpretation.plot(
    siamcat, 
    fn.plot = 'interpretationMND.pdf',
    color.scheme = "BrBG",
    consens.thres = 0.5,
    heatmap.type = "zscore",
    limits = c(-3, 3), detect.lim = 1e-06,
    max.show = 50, prompt=TRUE, verbose = 1)

@@

check.confounders(siamcat, fn.plot = './confoundersMND.pdf')

@@

print(length(unique(pseq$subjectID))) 
print(nrow(pseq))

print(length(unique(pseq@sam_data[["subjectID"]])))
#6784 total rows

@@

pseq.s<- pseq %>%
  select(sampleID, subjectID, study_condition, disease_subtype, 
disease, age, country, number_reads, median_read_length, BMI, mnd) %>%
  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
  mutate(visit=str_remove(visit, '_')) %>% 
  mutate(visit=as.numeric(visit)) %>%
  mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
  group_by(subjectID) %>% 
  filter(visit==min(visit)) %>%
  ungroup() %>%
  mutate(Sample_ID=sampleID) %>%
  mutate(Group=case_when(mnd=='LOW'~'CTR'))

>>>>

pseq.c <- core(pseq, detection = 0.1/100, prevalence = 50/100)

#pseq.subset <- subset_samples(pseq.c, !is.na(mnd) & disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" & antibiotics_current_use != "yes")

pseq.subset <- subset_samples(pseq.c, disease == "healthy" & age_category != "child" & age_category != "newborn" & age_category != "schoolage" )

pseq.c <- microbiome::transform(pseq.subset, "compositional")

@@

label <- create.label(meta=sample_data(pseq.c),
    label = "mnd",
    case = c("HIGH", "MEDIUM"), control = "LOW")

#FRIST WAY
# siamcat <- siamcat(phyloseq=phyloseq, label=label)
siamcat <- siamcat(phyloseq=pseq.c, label=label)

#SECOND WAY:
# siamcat <- siamcat(feat=feat, label=label, meta=meta)
# siamcat <- siamcat(feat=otu_table, label=label, meta=sample_data(pseq))
# Error in validObject(.Object) : invalid class “phyloseq” object: 
 # An otu_table is required for most analysis / graphics in the phyloseq-package

@@

## Feature Filtering
Now, we can filter feature with low overall abundance and prevalence.
Since we have quite a lot of microbial species in the dataset at the moment, we can perform unsupervised feature selection using the function filter.features.
```{r}

@@

#### Association Testing ####
Checks for associations of single species with disease suing nonparametric Wilcoxon test.

The check.assocation function calculates the significance of enrichment and metrics of association (such as generalized fold change and single-feautre AUROC).

siamcat <- check.associations(
    siamcat,
    sort.by = 'fc',
    fn.plot = 'assocMND_healthy.pdf',
    alpha = 0.05,
    mult.corr = "fdr",
    detect.lim = 10 ^-6,
    plot.type = "quantile.box",
    panels = c("fc", "prevalence", "auroc"))

@@

check.confounders(siamcat,
    fn.plot = 'conf_checkMND_healthy.pdf', meta.in = c("mnd", "gender", "age_category"))

@@

Getting all data on the same scale: if the scales for different features are wildly different, this can have a knock-on effect on your ability to learn (depending on what methods you're using to do it). Ensuring standardised feature values implicitly weights all features equally in their representation.

@@

#library(randomForest)

@@

#### Interpretation Plot
Displays importance of individual features in the classification model

@@

model.interpretation.plot(
    siamcat, 
    fn.plot = 'interpretationMND_healthy.pdf',
    color.scheme = "BrBG",
    consens.thres = 0.5,
    heatmap.type = "zscore",
    limits = c(-3, 3), detect.lim = 1e-06,
    max.show = 50, prompt=TRUE, verbose = 1)

model.interpretation.plot(
    siamcat, 
    fn.plot = 'interpretationfcMND.pdf',
   color.scheme = "BrBG",
    consens.thres = 0.5,
    heatmap.type = "fc",
    limits = c(-3, 3), detect.lim = 1e-06,
    max.show = 50, prompt=TRUE, verbose = 3)

@@

#check.confounders(siamcat, fn.plot = './confoundersMND.pdf')

@@

#print(length(unique(pseq$subjectID))) 
#print(nrow(pseq))

#print(length(unique(pseq@sam_data[["subjectID"]])))
#6784 total rows

@@

#pseq.s<- pseq %>%
  #select(sampleID, subjectID, study_condition, disease_subtype, 
#disease, age, country, number_reads, median_read_length, BMI, mnd) %>%
#  mutate(visit=str_extract(sampleID, '_[0-9]+$')) %>%
#  mutate(visit=str_remove(visit, '_')) %>% 
#  mutate(visit=as.numeric(visit)) %>%
 # mutate(visit=case_when(is.na(visit)~0, TRUE~visit)) %>%
  #group_by(subjectID) %>% 
  #filter(visit==min(visit)) %>%
  #ungroup() %>%
  #mutate(Sample_ID=sampleID) %>%
  #mutate(Group=case_when(mnd=='LOW'~'CTR'))
